---
title: 'Compute Centered and Scaled Genotype Matrix: W'
author: "Izel Fourie Sørensen"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/Izel/Dropbox/qgg-usersguide")
```

```{r, echo=FALSE}
rm(list=ls(all=TRUE))
```


Here we present two approaches for preparing the centered and scaled genotype matrix, $\mathbf{W}$. The genotype data prepared earlier $\mathbf{W}$ is computed based on the genotype data prepared earlier in the qgg user guide.



### Approach 1

The additive genomic relationship matrix $\mathbf{G}$ (VanRaden PM. 2008. J Dairy Sci. 91:4414-4423) is constructed using all genetic markers as follows: $\mathbf{G=WW}'/m$, where $\mathbf{W}$ is the centered and scaled genotype matrix, and $m$ is the total number of markers. Each column vector of $\mathbf{W}$ was calculated as follows: $\mathbf{w}_i=(m_i- 2p_i)/ \sqrt{(2p_i(1-p_i)}$, where $p_i$ is the minor allele frequency of the i^th^ genetic marker and $\mathbf{m}_i$ is the i^th^ column vector of the allele count matrix, $\mathbf{M}$, which contains the genotypes coded as 0, 1 or 2 counting the number of minor allele. (For the nearly homozygous DGRP lines the genotypes are coded as 0 or 2.)

Load the edited SNP genotype data file, `snpGE.Rdata`, created previously in the qgg user guide. The genotype data frame, `snpG`, is loaded  directly into the object `W` by the `readRDS()` function. 
```{r, eval=FALSE}
W <- readRDS(file="./genotypes/snpGE.rds")
```

Count the number of minor alleleles (`nMinor`) and total number of alleles (`nAlleles`) for each SNP. Actually for `nAlleles` we count the number of genotypes that were measured per row/SNP (thus not  "NA's"). This corresponds to counting one allele per genotype. Therefore the total number of alleles are 2 times `nAlleles`.
```{r, eval=FALSE}
nMinor <- rowSums(W, na.rm=TRUE)
nAlleles <- rowSums(!is.na(W))
```    
    
Compute minor allele frequencies:
```{r, eval=FALSE}
p<-nMinor/(2*nAlleles)
  min(p)
  max(p)
```

Center and scale $\mathbf{W}$ using the observed allele frequencies:
```{r, eval=FALSE}
for ( i in 1:205) {
  W[,i] <- (W[,i]-2*p)/sqrt(2*p*(1-p)) 
  isNA <- is.na(W[,i])
  W[isNA,i] <- 0
}
W <- t(W)
```

Save centered and scaled $\mathbf{W}$ calculated with approach 1 as `dgrp2_W1.Rdata`.
```{r, eval=FALSE}
save(W, file="./genotypes/dgrp2_W1.Rdata")
```



###Approach 2


In this approach the columns in $\mathbf{W}$ is scaled using the `scale()` function whose default method centers and/or scales the columns of a numeric matrix.

```{r, eval=FALSE}
rm(list=ls(all=TRUE))
```


Load the edited SNP genotype data file:
```{r, eval=FALSE}
W <- readRDS(file="./genotypes/snpGE.rds")
```

$\mathbf{W}$ is transposed because the `scale()` function by default scales the *columns* of a numeric matrix. 
```{r, eval=FALSE}
W <- t(W)
W <- scale(W)
dim(W)
```

Set missing values equal to 0.
```{r, eval=FALSE}
for ( i in 1:205) {
  isNA <- is.na(W[i,])
  W[i, isNA] <- 0
}
```

Save centered and scaled $\mathbf{W}$ calculated with approach 2 as `dgrp2_W2.Rdata`.
```{r, eval=FALSE}
save(W, file="./genotypes/dgrp2_W2.Rdata")
```

