---
title: "Prepare Phenotype and Covariate Data"
author: "Izel Fourie Sorensen"
date: "April 3, 2018"
output: html_document
---



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "C:/Users/Izel/Dropbox/qgg-usersguide")
```

```{r, echo=FALSE}
rm(list=ls(all=TRUE))
```


In this script we prepare a phenotype and covariate data frame to be used in downstream genomic analyses. As an example of phenotype data we are using the phenotype "resistance to starvation" from the *Drosophila melanogaster* Genetic Reference Panel (DGRP). The data is available at <http://dgrp2.gnets.ncsu.edu/data.html> under "Phenotype files", "Mackay, et al., Nature, 2012". Data for both males and females are used. Inversion status (chromosomal inversions) and Wolbachia (*Wolbachia* infection status) can be found under the heading "Other useful files" at the bottom of the page.   

We use the plyr, dplyr and tidyr packages for editing data. The readxl package is used for reading .xlsx files. Install these packages as follows:
```{r, eval=FALSE}
#install.packages("plyr")
#install.packages("dplyr")
#install.packages("tidyr")
#install.packages("readxl")
```
  
  
```{r, message=FALSE}
library(plyr)
library(dplyr)
library(tidyr)
library(readxl)
```
<br/> 
  
#### Download phenotype and covariate data

```{r}
# Female
download.file("http://dgrp2.gnets.ncsu.edu/data/website/starvation.female.csv", 
              destfile = "./data/starvation.female.csv")

# Male
download.file("http://dgrp2.gnets.ncsu.edu/data/website/starvation.male.csv", 
              destfile = "./data/starvation.male.csv")

# Inverstion status
download.file("http://dgrp2.gnets.ncsu.edu/data/website/inversion.xlsx", mode = "wb", 
              destfile = "./data/inversion.xlsx")

# Wolbachia
download.file("http://dgrp2.gnets.ncsu.edu/data/website/wolbachia.xlsx", mode = "wb", 
              destfile = "./data/wolbachia.xlsx")
```

#### Read and edit phenotype data

Read female data.
```{r}
starF <- read.csv(file="./data/starvation.female.csv", header = FALSE)
head(starF)
dim(starF)
```
\hfill\break  
Read male data.
```{r}
starM <- read.csv(file="./data/starvation.male.csv", header = FALSE)
head(starM)
dim(starM)
```
\hfill\break  
Give column names. "L" = lines, "F" = female, "M" = male.
```{r}
colnames(starF) <- c("L", "F")
colnames(starM) <- c("L", "M")
```
  
In dplyr a data frame has to be converted to a tibble (`tbl`). Convert `starF` and `starM` to tibbles.
```{r}
starF <- tbl_df(starF)
starM <- tbl_df(starM)
```

Look at the tibbles
```{r}
starF
starM
```
\hfill\break  
Join the tibbles for males and females.
```{r}
starMF <- left_join(starM, starF, by= "L")
starMF
```
\hfill\break  
Create a column for sex information and a column for the phenotype (y), in this case resistance to starvation.
```{r}
starv <- gather(starMF, sex, y, -L)
head(starv)
head(starv, 3)
```
\hfill\break  
Remove prefix "line_" from the contents of the "L" column.
```{r}
starv$L <- gsub("line_", "", starv$L, fixed = TRUE)
head(starv$L)
```
<br>
  
#### Read and edit Inversion status
  
Abbreviations used are: INV = inversion karyotype (homozygous), INV / ST = heterozygote for the inversion and ST = standard configuration in a homozygous form.
```{r}
inv <- read_excel("./data/inversion.xlsx",
                   sheet = 1, col_names = TRUE)

head(inv)
dim(inv)
```
\hfill\break  
Remove the "DGRP_" prefix from the contents of the "DGRP Line" column. Save column names of `Inv` as a vector `Invcols`. Then edit the column names in the `Invcols` vector: "DGRP Line" becomes "L" and the brackets "()" in the inversion names are changed to underscores.
```{r}
inv$`DGRP Line` <- gsub("DGRP_", "", inv$`DGRP Line`, fixed=TRUE)
head(inv, 3)

invcols <- colnames(inv)    
invcols[1] <- "L"
invcols[2:17] <- gsub("(", "", invcols[2:17], fixed = TRUE)
invcols[2:17] <- gsub(")", "", invcols[2:17], fixed = TRUE)
colnames(inv) <- invcols
inv
```

  
#### Read and edit *Wolbachia* status

```{r}
wo <- read_excel("./data/wolbachia.xlsx",
                   sheet = 1, col_names = TRUE)
wo
dim(wo)
```
  
Change column names of `wo`
```{r}
colnames(wo) <- c("L", "wo")
wo$L <- gsub("DGRP__", "", wo$L, fixed=TRUE)
wo
```
 
#### Create final data frame 
Merge phenotype data with inversion status and *Wolbachia* infection status.
```{r}
starvInv <- left_join(starv, inv, by= "L")
starvInv

starvIW <- left_join(starvInv, wo, by="L")
starvIW[1:5,15:20]
```

An example of how one can look at the data. Here we show a summary (in table form) of the first three inversions.
```{r}
apply(starvIW[,4:19], 2, table)[1:3]
```

Convert the tibble to a data frame and save the edited phenotype data.
```{r}
starv <- as.data.frame(starvIW)
save(starv, file="./phenotypes/starv_inv_wo.Rdata")
```

